Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-06-18 Compare_053014_statsmed_revisions.tex
Updated tables and analysis using KM estimate of integrated STG and new simulation results

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T. The estimated marginal distribution of event times $\hat{F}(T)$ can be calculated in a number of ways, by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5 or Kaplan-Meier estimator. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
\begin{tabular}{lccccccc}
  \hline
&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$&$STG(3)^{*}=0.23$&$STG(3)^{*}=0.51$\\ 
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
 &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
 &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
 &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
Bias $STG.i$ & 0.8 & 0.065 (0.079)& 0.021 (0.081)& 0.023 (0.051)& 0.003 (0.067) & 0.009 (0.075) & 0.041 (0.074) \\
 & 0.5 & 0.059 (0.099) & 0.052 (0.109) & 0.006 (0.053)& 0.007 (0.076)& 0.076 (0.224) & 0.024 (0.153) \\
Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   \hline
\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$. *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, 3.5)$ years \label{IPow}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}=0.47$& $\widetilde{STG}^*=0.23$  & $\widetilde{STG}^*=0.52$\\
  \hline
$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 75.6 & 24.6 & 56 \\ 
&  0.5  & 7.2 & 37.6 & 91.8 & 93.9 & 97.4 \\ 
 $\widetilde{STG}=0.25$ & 0.8 &  & 4 & 63.6 & 47.2 & 37.6 \\ 
&  0.5 &  & 4 & 78.8 & 80 & 94.8 \\ 
$\widetilde{STG}=0.47$ & 0.8 &  &  & 6 & 7.5 & 8.8 \\ 
 & 0.5 &  &  & 6 & 2.8 & 16.2 \\ 
$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 24.2 \\ 
&  (0.5, 0.5) &  &  &  & 6.3 & 17.2 \\ 
$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 1.6 \\ 
 & (0.5, 0.5) &  &  &  &  & 6 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001. The P-value for testing H0$_1$ for EGFR is 0.19. The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055. The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of 0.018. Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone. The estimated $\widehat{\widetilde{STG}}$ is 0.2 for HBA1C alone and 0.15 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.202. Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.

The P-value is 0.057 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR. This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR. As the difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS is so small, it is not surprising that the P-value for the null that their difference is equal to zero is 0.99.

>>>>

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T. The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
\begin{tabular}{lccccccc}
  \hline
&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$&$STG(3)^{*}=0.23$&$STG(3)^{*}=0.51$\\ 
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
 &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
 &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
 &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & 0.023 (0.093) & 0.032 (0.054) \\
 & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.022 (0.095) & 0.069 (0.092) \\
Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   \hline
\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$. *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, \infty)$ years \label{IPow}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}=0.47$& $\widetilde{STG}^*=0.23$  & $\widetilde{STG}^*=0.52$\\
  \hline
$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 99.6 & 7.8 & 99.2 \\ 
&  0.5  & 5.7 & 17.8 & 99.8 & 9.2 & 99.4 \\ 
 $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 91.6 & 6.6 & 83.6 \\ 
&  0.5 &  & 5.5 & 95.2 & 7.1 & 89.8 \\ 
$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 82.6 & 8.8 \\ 
 & 0.5 &  &  & 5.2 & 81.9 & 16.2 \\ 
$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 75.8 \\ 
&  (0.5, 0.5) &  &  &  & 6.3 & 78.0 \\ 
$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.6 \\ 
 & (0.5, 0.5) &  &  &  &  & 5.9 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001. The P-value for testing H0$_1$ for EGFR is 0.19. The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055. The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of 0.018. Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone. The estimated $\widehat{\widetilde{STG}}$ is 0.31 for HBA1C alone and 0.21 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.35. Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.

The P-value is 0.058 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR. This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR. The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS gives a p-value of 0.18.
